Robust Trajectory Optimization on a Quantum System

### 2021-01-26 paper/3fluxonium.tex
revised based on dave comments

\section{QOC for the Fluxonium \label{sec:fluxonium}}
In the following, we optimize quantum gates
(unitary transformations) for the fluxonium qubit.
The fluxonium is a promising building block
for quantum computers, and the accurate
two-level approximation of its Hamiltonian makes
QOC on a classical computer inexpensive.
In the two-level
approximation, the Hamiltonian takes the form:
\begin{align}
  H/h &= f_{q} \frac{\sigma_{z}}{2} + a(t) \frac{\sigma_{x}}{2}
  \label{eq:hamiltonian}
\end{align}
where $f_{q}$ is the qubit frequency at the flux frustration point,
$a(t)$ is the flux-drive amplitude, $h$ is Planck's constant, and $\sigma_{z}, \sigma_{x}$
are Pauli matrices. We optimize $X/2$, $Y/2$, and
$Z/2$ gates for the fluxonium presented in \cite{zhang2020universal},
and compare them to the analytically constructed gates for
that device.

First, we outline the constraints for the fluxonium gate problem.
We formulate this problem as a multi-state-transfer problem.
The initial conditions on
the states are $\ket{\psi^{0}_{1}} = \ket{0}$, $\ket{\psi^{1}_{1}} = \ket{1}$
\eqref{eq:istate_con}
where the superscript is an index $i \in \{0, 1\}$,
and the subscript indicates the first knot point $k = 1$.
The states at the final knot point are constrained to be
the image of the initial states under the desired gate $U$,
$\ket{\psi^{i}_{N}} = U \ket{\psi^{i}_{1}} \ \forall \ i$
\eqref{eq:tstate_con}.
%% This constraint is sensitive to global phase
%% differences between the states.
%% , but it is computationally efficient
%% because the Jacobian and Hessian
%% of the corresponding constraint function are diagonal.
%% We could have analogously chosen a metric which is insensitive
%% to global phases, such as the fidelity of the states.
Furthermore, we impose the normalization constraint
${\lvert \braket{\psi^{i}_{k}}{\psi^{i}_{k}} \rvert}^{2} = 1 \ \forall \ i,k$
\eqref{eq:statenorm_con}
to ensure the solver does not take advantge of discretization errors in numerical integration.
To refer to the discrete moments the flux amplitude, we introduce the notation
$\int^{t_{k}}_{t_{1}} a(t) \ \mathrm{d}t \equiv \int_{t} a_{k}$,
$a(t) \lvert_{t = t_{k}} \equiv a_{k}$,
$\pdv*[n]{a(t)}{t} \lvert_{t = t_{k}} \equiv \partial^{n}_{t} a_{k}$.
We impose the zero net flux constraint $\int_{t} a_{N} = 0$
\eqref{eq:znf_con}
which mitigates the hysteresis ubiquitous in flux-bias lines
\cite{battistel2019fast, krantz2019quantum, zhang2020universal}.
The flux amplitude is constrained by $\lvert a_{k} \rvert \leq 0.5 \ \textrm{GHz} \ \forall \ k$
\eqref{eq:amp_con}
to ensure the two-level approximation \eqref{eq:hamiltonian} remains valid.
We also enforce the boundary condition $a_{1} = a_{N} = 0$ \eqref{eq:bound_con}
so the gates may be concatenated arbitrarily. Additionally,
we have the initial condition $\int_{t} a_{1} = \partial_{t} a_{1} = 0$
\eqref{eq:ic_con}.

@@

  \begin{subfigure}{.315\textwidth}
    \includegraphics[width=\linewidth]{assets/f1a.png}
    \caption{\label{fig:longitudea}}
  \end{subfigure}\hfill
  \begin{subfigure}{.23\textwidth}
    \includegraphics[width=\linewidth]{assets/f1b.png}
    \caption{\label{fig:longitudeb}}
  \end{subfigure}\hfill
  \begin{subfigure}{.4\textwidth}
    \includegraphics[width=\linewidth]{assets/f1c.png}
    \caption{\label{fig:longitudec}}
  \end{subfigure}
  \caption{
    (a) Numerically optimized gates (dark blue)
    and analytically optimized gates (light pink).
    (b) $T_{1}$ interpolation function used in optimization. Markers
    denote the time-averaged, absolute amplitude of each gate.
    (c) Cumulative gate errors due to depolarization as a function of the
    number of gates applied.
    Gate errors for the numerically optimized $Z/2$ and $Y/2$ gates
    are indistinguishable. Inset shows the cumulative gate errors
    plotted on a log scale for small gate counts.
  }
  \label{fig:longitude}
\end{figure*}

Next, we outline the cost functions for the fluxonium gate problem.
The augmented state and augmented control are:
\begin{equation}
  \begingroup
  \renewcommand*{\arraystretch}{1.3}
  x_{k} = \begin{bmatrix} \ket{\psi^{0}_{k}} \\ \ket{\psi^{1}_{k}}
    \\ \int_{t} a_{k} \\ a_{k} \\ \partial_{t} a_{k} \end{bmatrix} \quad
  u_{k} = \begin{bmatrix} \partial^{2}_{t} a_{k} \end{bmatrix}
  \endgroup
  \label{eq:astatecontrols}
\end{equation}
The cost function at each knot point is
$\ell_{k}(x_{k}, u_{k}) = (x_{k} - x_{f})^{T} Q_{k} (x_{k} - x_{f}) + u^{T}_{k} R_{k} u_{k}$
where $Q_{k}$ and $R_{k}$ are matrices we supply, and
$x_{f}$ is given by the constraints we have imposed on
$\ket{\psi^{i}_{N}}$, $\int_{t} a_{N}$, and $a_{N}$ in addition to
$\partial_{t} a_{f} = 0$. We choose $Q_{k}$ to be diagonal
so the cost function incentivizes the achievement of the constraints on
$\ket{\psi^{i}_{N}}$, $\int_{t} a_{N}$, and $a_{N}$,
and penalizes the norm of $\partial_{t} a_{k}$ and $\partial^{2}_{t} a_{k}$,
mitigating AWG ringing due to high-frequency transitions.
In the discrete dynamics function for this problem \eqref{eq:dyn_con},
we integrate the states according to the TDSE \eqref{eq:tdse} and the
fluxonium Hamiltonian \eqref{eq:hamiltonian} and integrate
the second-derivative of the flux amplitude to obtain
the first-derivative, proportional, and integral moments.
Stated succinctly, the optimization problem takes the form:
\begin{mini!}[2] 
  {x_{1:N}, u_{1:N\text{-}1}}{\sum_{k=1}^N {(x_k\text{-}x_f)}^{T} Q_k (x_k\text{-}x_{f})
    + \sum_{k=1}^{N-1} {u_k}^{T} R_k u_{k}}{}{} \label{eq:costfun}
  \addConstraint{x_{k+1}}{= f(x_k, u_k)}  \label{eq:dyn_con}
  \addConstraint{\ket{\psi^{0}_{1}} = \ket{0}, \ket{\psi^{1}_{1}} = \ket{1}} \label{eq:istate_con}
  \addConstraint{\ket{\psi^{i}_{N}} = U \ket{\psi^{i}_{1}}
    \ \forall \  i} \label{eq:tstate_con}
  \addConstraint{{\lvert \braket{\psi^{i}_{k}}{\psi^{i}_{k}}
      \rvert}^{2} = 1 \ \forall \ i, k} \label{eq:statenorm_con}
  \addConstraint{{\textstyle \int_{t}} a_{N} = 0} \label{eq:znf_con}
  \addConstraint{|a_{k}| \leq 0.5 \ \textrm{GHz} \ \forall \ k} \label{eq:amp_con}
  \addConstraint{a_{1} = a_{N} = 0} \label{eq:bound_con}
  \addConstraint{{\textstyle \int_{t}} a_{1} = \partial_{t} a_{1} = 0} \label{eq:ic_con}
\end{mini!}

>>>>

\section{QOC for the Fluxonium \label{sec:fluxonium}}
In the following, we optimize quantum gates
(unitary transformations) for the fluxonium qubit.
The fluxonium is a promising building block
for quantum computers, and the accurate
two-level approximation of its Hamiltonian makes
QOC on a classical computer inexpensive.
To high accuracy, we approximate the Hamiltonian near the flux frustration
as a spin-$1/2$ system:
\begin{align}
  H/h &= f_{q} \frac{\sigma_{z}}{2} + a(t) \frac{\sigma_{x}}{2},
  \label{eq:hamiltonian}
\end{align}
where $f_{q}$ is the qubit frequency at the flux frustration point,
$a(t)$ is the flux bias, $h$ is Planck's constant, and $\sigma_{z}, \sigma_{x}$
are Pauli matrices. Although we use an approximation for the Hamiltonian,
our noise model is realistic and considers the full Hilbert space. We optimize $X/2$,
$Y/2$, and $Z/2$ gates for the fluxonium presented in \cite{zhang2020universal},
and compare them to the analytically constructed gates for
that device.

First, we outline the constraints for the fluxonium gate problem.
We formulate this problem as a multi-state-transfer problem.
The initial conditions on
the states are $\ket{\psi^{0}_{1}} = \ket{0}$, $\ket{\psi^{1}_{1}} = \ket{1}$
\eqref{eq:istate_con}
where the superscript is an index $i \in \{0, 1\}$,
and the subscript indicates the first knot point $k = 1$.
The states at the final knot point are constrained to be
the image of the initial states under the desired gate $U$,
$\ket{\psi^{i}_{N}} = U \ket{\psi^{i}_{1}} \ \forall \ i$
\eqref{eq:tstate_con}.
Furthermore, we impose the normalization constraint
${\lvert \braket{\psi^{i}_{k}}{\psi^{i}_{k}} \rvert}^{2} = 1 \ \forall \ i,k$
\eqref{eq:statenorm_con}
to ensure the solver does not take advantage of discretization errors in numerical integration.
To refer to the discrete moments the flux, we introduce the notation
$\int^{t_{k}}_{t_{1}} a(t) \ \mathrm{d}t \equiv \int_{t} a_{k}$,
$a(t = t_{k}) \equiv a_{k}$,
$\pdv*[n]{a(t)}{t} \lvert_{t = t_{k}} \equiv \partial^{n}_{t} a_{k}$.
We impose the zero net flux constraint $\int_{t} a_{N} = 0$
\eqref{eq:znf_con}
which mitigates the inductive drift ubiquitous in flux-bias lines
\cite{battistel2019fast, krantz2019quantum, zhang2020universal}.
The flux is constrained by $\lvert a_{k} \rvert \leq 0.5 \ \textrm{GHz} \ \forall \ k$
\eqref{eq:amp_con}
to ensure the two-level approximation \eqref{eq:hamiltonian} remains valid.
We also enforce the boundary condition $a_{1} = a_{N} = 0$ \eqref{eq:bound_con}
so the gates may be concatenated arbitrarily. Additionally,
we have the initial condition $\int_{t} a_{1} = \partial_{t} a_{1} = 0$
\eqref{eq:ic_con}.

@@

  \begin{subfigure}{.315\textwidth}
    \includegraphics[width=\linewidth]{assets/f1a.png}
    \caption{\label{fig:longitudea}}
  \end{subfigure}\hfill
  \begin{subfigure}{.23\textwidth}
    \includegraphics[width=\linewidth]{assets/f1b.png}
    \caption{\label{fig:longitudeb}}
  \end{subfigure}\hfill
  \begin{subfigure}{.4\textwidth}
    \includegraphics[width=\linewidth]{assets/f1c.png}
    \caption{\label{fig:longitudec}}
  \end{subfigure}
  \caption{
    (a) Flux pulses for the numerical gates (dark blue)
    and the analytic gates (light pink).
    (b) $T_{1}$ interpolation function used in optimization. Circle markers
    indicate measured $T_{1}$ times. Non-circle markers
    indicate the time-averaged, absolute amplitude of each flux pulse.
    (c) Cumulative gate errors due to depolarization as a function of the
    number of gates applied.
    Cumulative gate errors for the numerical $Z/2$ and $Y/2$ gates
    are indistinguishable. Inset shows log-scaled cumulative gate errors
    for small gate counts.
  }
  \label{fig:longitude}
\end{figure*}

Next, we introduce the augmented control and augmented state:
\begin{equation}
  \begingroup
  \renewcommand*{\arraystretch}{1.3}
  u_{k} = \begin{bmatrix} \partial^{2}_{t} a_{k} \end{bmatrix}, \quad
  x_{k} = \begin{bmatrix} \ket{\psi^{0}_{k}} \\ \ket{\psi^{1}_{k}}
    \\ \int_{t} a_{k} \\ a_{k} \\ \partial_{t} a_{k} \end{bmatrix}.
  \endgroup
  \label{eq:astatecontrols}
\end{equation}
The variables in the augmented state are derived from
the decision variable in the augmented
control via coupled, first-order differential equations in the 
the discrete dynamics function \eqref{eq:dyn_con}.
We integrate the states according to the TDSE \eqref{eq:tdse} and the
fluxonium Hamiltonian \eqref{eq:hamiltonian} and integrate
the second-derivative of the flux amplitude to obtain
the first-derivative, proportional, and integral moments.
The ALTRO implementation we use does not currently
support complex numbers, so we repesent the states
in the isomorphism $\mathcal{H}(\mathbb{C}^{n})
\cong \mathcal{H}(\mathbb{R}^{2n})$ given in \cite{leung2017speedup},
\begin{equation}
  H \ket{\psi} \cong \begin{bmatrix} H_{\textrm{re}} & -H_{\textrm{im}}
    \\ H_{\textrm{im}} & H_{\textrm{re}}\end{bmatrix}
  \begin{bmatrix} \ket{\psi}_{\textrm{re}} \\ \ket{\psi}_{\textrm{im}}\end{bmatrix}.
  \label{eq:isomorphism}
\end{equation}
The cost function at each knot point is
$\ell_{k}(x_{k}, u_{k}) = (x_{k} - x_{f})^{T} Q_{k} (x_{k} - x_{f}) + u^{T}_{k} R_{k} u_{k}$
where $Q_{k}$ and $R_{k}$ are diagonal matrices we supply. The $Q_{k}$ term
penalizes deviations from the final augmented state $x_{f}$,
which is given by the constraints we have imposed on
$\ket*{\psi^{i}_{N}}$, $\int_{t} a_{N}$, and $a_{N}$ in addition to
$\partial_{t} a_{f} = 0$. The $R_{k}$ term penalizes the norm of $\partial^{2}_{t} a_{k}$,
mitigating AWG ringing due to high-frequency transitions.
Stated succinctly, the optimization problem takes the form:
\begin{mini!}[2] 
  {x_{1:N}, u_{1:N\text{-}1}}{\sum_{k=1}^N {(x_k\text{-}x_f)}^{T} Q_k (x_k\text{-}x_{f})
    + \sum_{k=1}^{N-1} {u_k}^{T} R_k u_{k}}{}{} \label{eq:costfun}
  \addConstraint{x_{k+1}}{= f(x_k, u_k)}  \label{eq:dyn_con}
  \addConstraint{\ket{\psi^{0}_{1}} = \ket{0}, \ket{\psi^{1}_{1}} = \ket{1}} \label{eq:istate_con}
  \addConstraint{\ket{\psi^{i}_{N}} = U \ket{\psi^{i}_{1}}
    \ \forall \  i} \label{eq:tstate_con}
  \addConstraint{{\lvert \braket{\psi^{i}_{k}}{\psi^{i}_{k}}
      \rvert}^{2} = 1 \ \forall \ i, k} \label{eq:statenorm_con}
  \addConstraint{{\textstyle \int_{t}} a_{N} = 0} \label{eq:znf_con}
  \addConstraint{|a_{k}| \leq 0.5 \ \textrm{GHz} \ \forall \ k} \label{eq:amp_con}
  \addConstraint{a_{1} = a_{N} = 0} \label{eq:bound_con}
  \addConstraint{{\textstyle \int_{t}} a_{1} = \partial_{t} a_{1} = 0}. \label{eq:ic_con}
\end{mini!}

### 2021-01-26 paper/4longitude.tex
revised based on dave comments

\section{Depolarization Awareness \label{sec:longitude}}
The strength of depolarization varies
with the control parameters in a range of superconducting circuit platforms,
including the fluxonium. In this section, we outline a method
for tuning the flux amplitude to mitigate depolarization.
Previous work has modeled the gate error due to depolarization
by evolving density matrices under a master
equation \cite{rembold2020introduction, schulteherbruggen2011optimal}
or evolving a large number of states in a quantum trajectory approach
\cite{abdelhafez2019gradient}.
We avoid the increase in computational complexity required for these
techniques by
penalizing the integrated rate (probability) of depolarization in optimization.
Using this probability as a proxy for the gate error incurred
is reasonable because losses due to depolarization increase monotonically
in time. Additionally, for a constant $T_{1}$ time--the $1/e$ depolarization
time for the qubit--a shorter gate duration
would favor a lower depolarization probability. We allow
the optimizer to tune the gate time to minimize the
depolarization probability.

The depolarization probability is given by,
\begin{equation}
  P_{1}(t) = \int_{0}^{t} T_{1}^{-1}(a(t^{\prime})) dt^{\prime}
\end{equation}
This value is appended to the augmented state \eqref{eq:astatecontrols}
and its norm is penalized in the objective \eqref{eq:costfun} by setting
the corresponding element of the final augmented state $x_{f}$ to $0$.
$T_{1}(a_{k})$ is obtained at each knot point by evaluating
a spline fit to experimental data of the form $\{(a, T_{1})\}$,
see Figure \ref{fig:longitudeb}.
It is also possible to fit a spline to theoretically obtained data.
However, $T_{1}$ values are known to fluctuate greatly
with laboratory temperatures \cite{klimov2018fluctuations}.
Interpolating $T_{1}$ from experimental data
increases the fridge truth of the simulation.

@@

We compare the performance of the $X/2$, $Y/2$, and $Z/2$ gates obtained with
our numerical method to the analytic gates. We
use the Lindblad master equation to simulate $T_{1}$ dissipation for successive
gate applications, and compute the cumulative gate error
after each application, see Appendix \ref{appendix:longitude}.
The flux amplitudes for the numerical gates are similar; the
waveforms are approximately periodic
with amplitudes $\sim 0.2 \textrm{GHz}$, see Figure \ref{fig:longitudea}. \todo{Floquet}.
Their gate times are greater
than their analytic counterparts, but they
reach higher amplitudes and therefore higher $T_{1}$ times.
Although the analytic and numerical gates attain single gate errors sufficient for
quantum error correction ($< 10^{-4}$), which are required for fault-tolerant quantum computing
\cite{aharonov2008fault, fowler2009high, gottesman1997stabilizer},
the single gate errors for the numerical
gates are a factor of $5$ less
than those for the analytic gates, commensurate to the
depolarization probability reductions, see Appendix \ref{appendix:longitude}.
The gate error reported in this text is the infidelity
of the evolved state and the target state averaged over 1000 pseudo-randomly
generated initial states.
The numerical $Z/2$ and $Y/2$ gates produce similar
results in the concatenated gate application comparison,
yielding cumulative gate errors below $8 \cdot 10^{-3}$
over $2000$ gate applications $\sim 40 \mu\textrm{s}$, see Figure \ref{fig:longitudec}.
The cumulative gate error for the numerical $X/2$ gate is
$1.7 \cdot 10^{-2}$ over $2000$ gate applications $\sim 124 \mu\textrm{s}$.
The low cumulative gate errors for high gate counts produced
by the numerical gates are critical for
noisy, intermediate-scale quantum (NISQ) applications.
These improvements are significant for the realistic constraints we have imposed
on the gates, and do not represent a fundamental limit to the optimization methods we have
employed.

>>>>

\section{Depolarization Mitigation\label{sec:longitude}}
In this section, we outline a method
for optimizing the flux to mitigate depolarization.
For many superconducting circuits, the 1/$e$ depolarization time ($T_{1}$) is independent of the control parameters,
so the fastest possible gate incurs the least depolarization error
\cite{schulteherbruggen2011optimal}.
For the fluxonium, however, $T_{1}$ is strongly dependent on the flux.
We enable the optimizer to trade longer gate times
for lower depolarizaiton times or vice-versa by making the gate time a decision variable.
Additionally, previous work has modeled the gate error due to depolarization
by evolving density matrices under a master
equation \cite{rembold2020introduction, schulteherbruggen2011optimal}
or evolving a large number of states in a quantum trajectory approach
\cite{abdelhafez2019gradient}.
We avoid the increase in computational complexity required for these
techniques by penalizing the integrated rate (probability) of depolarization in optimization.
Using this probability as a proxy for the gate error incurred
is reasonable because depolarization errors are incoherent and therefore
increase monotonically in time without interference.

The depolarization probability is given by,
\begin{equation}
  P_{1}(t) = \int_{0}^{t} T_{1}^{-1}(a(t^{\prime})) dt^{\prime}.
\end{equation}
This value is appended to the augmented state \eqref{eq:astatecontrols}
and its norm is penalized in the objective \eqref{eq:costfun} by setting
the corresponding element of the final augmented state to zero.
$T_{1}(a_{k})$ is obtained at each knot point by evaluating
a spline fit to experimental data of the form $\{(a, T_{1})\}$,
see Figure \ref{fig:longitudeb}.
It is also possible to fit a spline to theoretically obtained data.
However, experimental data often deviates from theoretical predictions
because it is difficult to account for all loss channels \cite{zhang2020universal},
and $T_{1}$ values often fluctuate on hour
time scales \cite{klimov2018fluctuations}.

@@

We analyze the effect of depolarization on
the $X/2$, $Y/2$, and $Z/2$ gates obtained with
our numerical method and the analytic gates. We
use the Lindblad master equation to simulate $T_{1}$ dissipation for successive
gate applications, and compute the cumulative gate error
after each application, see Appendix \ref{appendix:longitude}.
\todo{Jens, Danny, Ziwen, any idea what the correct way to simulate depolarization is
  when $T_{1}$ is time dependent?}.
The gate error reported in this text is the infidelity
of the evolved state and the target state averaged over 1000 pseudo-randomly
generated initial states.
The flux pulses for the numerical gates
are approximately periodic
with amplitudes $\sim 0.2 \textrm{GHz}$, see Figure \ref{fig:longitudea}.
They are reminiscent of the analytically determined Floquet operations
for a fluxonium described in \cite{huang2020engineering}
and realized in \cite{mundada2020floquet}.
The numerical gate times are greater
than the analytic gate times, but the
numerical flux pulses
spend more time at higher amplitudes, achieving higher $T_{1}$ times.
The single gate errors for both the analytic and numerical gates are
less than $10^{-4}$, which makes them sufficient for quantum error correction--a
prerequisite for fault-tolerant quantum
computing \cite{aharonov2008fault, knill2005quantum, gottesman1997stabilizer}.
However, the numerical gates achieve single gate errors
$\sim 5$ times less than those for the analytic gates,
which tracks closely with their relative improvement
on the depolarization probability metric, see Appendix \ref{appendix:longitude}.
This single gate error advantage corresponds to a significant reduction in error correction resources.
Furthermore, for successive gate applications, the gate error due to depolarization is approximately linear
in the gate count, which we expect for $t \ll T_{1}$, see Figure \ref{fig:longitudec}.
The gate error reduction for large gate counts is important for noisy, intermediate-scale quantum (NISQ)
applications. These improvements are significant for the constraints we have imposed on the gates,
and do not represent a fundamental limit to the optimization methods we have employed.
